Think deep for reef resilience

### 2025-08-16 paper/paper_draft.qmd
Tidy up time series discussion

Each Global Climate Model (GCM) shows similar relationships between reef clusters and
contextual analysis factors, thus further comparisons of these clusters for each analysis
factor are only shown for ACCESS-ESM1-5 (an intermediate Equilibrium Climate Sensitivity
scenario [@sfig-ecs]). Results for the additional GCMs can be found at Grier and Iwanaga
[-@grier2025]. However, there are substantial differences in the longevity of reef
resilience and maintenance of coral cover post 2030 between the GCMs, thus coral cover
timeseries are displayed for each GCM (displayed clustered at management-area scale to
preserve coarse spatial differences in cover trajectories) (@fig-gcm-cover-trajectories).

NorESM2-MM has the lowest Equilibrium Climate Sensitivity (2.5°C) and reefs exhibit
continuing recovery and decline dynamics up to 2075 under this possible climate future.
Recovery phases are predominantly seen in the ‘high’ cover cluster. The ‘medium’ cluster,
and to a lesser degree the ‘low’ cluster, also exhibit periods of recovery and increasing
proportions of coral cover up to 2045. Under NorESM2-MM the Far Northern management area has
lower CID distances between low and high cover clusters than in other management areas
[@tbl-cluster-cid], this is reflected in the low degree of cover maintenance
[@fig-gcm-cover-trajectories]. ACCESS-ESM1-5 has an intermediate ECS value of 3.87°C which
results in intermediate levels of coral cover maintenance (between those of lower and higher
ECS GCMs) [@fig-gcm-cover-trajectories], reflected in intermediate CID distance values
within most management areas [@tbl-cluster-cid]. In Townsville-Whitsunday and Far Northern
management areas, reefs in the high cluster are able to maintain coral cover levels around
20% to 2050 under ACCESS-ESM1-5. Under ACCESS-CM2 (4.66°C), EC-Earth3-Veg (4.30°C) and
GFDL-CM4 (4.1°C) GCMs cover trajectories are largely declining in all management areas,
reflected in lower CID distance values than other GCMs [@tbl-cluster-cid].

Under each GCM, in all management areas, low, medium and high cluster reefs have low levels
of distinction between DHW trajectories (@sfig-gcm-dhw-trajectories). The highest peak DHW
levels occur under GFDL-CM4 between 2045 and 2050, while greater increases in overall DHWs
time series occur under ACCESS-CM2 [@fig-methods-dhws].

![Reef area-relative coral cover time series clustered within each management area, for each
GCM analysed. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover
between 2030 and 2060 within each bioregion. ‘Medium’ clusters (orange) are clusters with
intermediate median area-relative cover between 2030 and 2060 within each bioregion. ‘High’
clusters (blue) are clusters with the highest median area-relative cover between 2030 and
2060 within each bioregion. Solid lines represent the cluster median time series and bands
represent the 95th percentile confidence intervals. Each management area is represented five
times (for each GCM), with each GCM displayed
horizontally.](../figs/GCM_cover_timeseries_plot.png){#fig-gcm-cover-trajectories
fig-align="center" width="14.82cm"}

Comparing across GCMs, it is common for reefs to switch cluster level, including from low to
high and high to low cluster transitions (@sfig-gcm-bioregion-cluster-assignment). Of 2429
reefs, 20% (503) of reefs have consistent management area scale cluster assignment across
all GCMs. Of the 503 consistently assigned reefs, 49% are within the Far Northern management
area, with 7, 13 and 31% occurring in the Cairns/Cooktown, Townsville/Whitsunday and
Mackay/Capricorn management areas respectively [@sfig-gcm-cluster-map]. Of the 503
consistently assigned reefs, 32, 7 and 61% are clustered into low, medium and high clusters
respectively. High assignment variability across GCMs is expected as spatial differences in
GCM DHW timeseries result in reefs being impacted at different relative magnitudes within
their management area across GCMs.

>>>>

There were substantial differences in the maintenance of coral cover post 2030 between the
GCMs. Figure [-@fig-gcm-cover-trajectories] illustrates the temporal behaviour of reefs by
their categorisation (Low, Medium and High). Results are displayed at the Management area
scale to illustrate the spatial differences in cover trajectories. Of the GCMs explored,
NorESM2-MM has the lowest Equilibrium Climate Sensitivity (2.5°C) and reefs exhibit
continuing recovery and decline dynamics up to 2075 under this hypothetical climate future.

Unsurprisingly, recovery phases are predominantly seen in the reefs in the High category,
with Medium and Low exhibiting lesser probability of such periods of recovery. Generally,
recovery and increases in coral cover are visible in simulated results up to 2045. Under
NorESM2-MM, the Far Northern Management area were found to have the most similar temporal
behaviour between Low and High categories compared to other Management areas
[@tbl-cluster-cid] due to the low degree of coral cover [@fig-gcm-cover-trajectories].

Intermediate levels of coral cover maintenance can be seen in ACCESS-ESM1-5 (ECS value of
3.87°C) when compared to the other GCMs (of higher and lower ECS values). Under these
intermediate conditions (in relative terms), there is potential for reefs in the High
category to maintain coral cover levels around 20% to 2050. The highest peak DHW levels
occur under GFDL-CM4 between 2045 and 2050, while greater increases in overall DHWs time
series occur under ACCESS-CM2 [@fig-methods-dhws]. Under conditions represented by high ECS
value GCMs - specifically, ACCESS-CM2 (4.66°C), EC-Earth3-Veg (4.30°C) and GFDL-CM4
(4.1°C) - all reefs are largely in decline across all management areas from 2055
[@tbl-cluster-cid].

![Time series of coral cover (relative to reef area) categorised within each Management area
(rows), for each GCM analysed (columns) based on temporal behaviour over 2030 and 2060. The
time series are categorised as "Low" (green), "Medium" (in orange), and "High" median cover
(in blue). Solid lines represent the cluster median time series and bands represent the 95th
percentile confidence
intervals.](../figs/GCM_cover_timeseries_plot.png){#fig-gcm-cover-trajectories
fig-align="center" width="14.82cm"}

**\[TI: Ben, Just noticed - we need to make the rows (Management areas) go from North to
South!\]**

Comparing across GCMs, it is common for reefs to switch between categories, including from
Low to High and vice versa (@sfig-gcm-bioregion-cluster-assignment). Of 2429 reefs, 20%
(503) of reefs have consistent category assignment across all GCMs. Of the 503 consistently
assigned reefs, 49% are within the Far Northern Management area, with 7%, 13% and 31%
occurring in the Cairns-Cooktown, Townsville-Whitsunday and Mackay-Capricorn Management
areas respectively [@sfig-gcm-cluster-map]. Of the 503 consistently assigned reefs, 32%, 7%
and 61% are clustered into Low, Medium and High clusters respectively. High assignment
variability across GCMs is expected as spatial differences in a given GCM's DHW trajectory
result in reefs being impacted at different relative magnitudes within each Management
areas.
